Left-linear rewriting in adhesive categories

### 2024-04-21 concur/fusione.tex
minima

\begin{figure}
  \begin{subfigure}{0.2\textwidth}
    {\small
      \xymatrix@C=16pt@R=1pt{
        & 0 \ar@{-}[dd] \ar@{-}[ddddddl] \ar@{-}[ddddddr] &  \\
        &  & \\
        & 1 \ar@{-}[dd] \ar@{-}[ddddl]  \ar@{-}[ddddr]   &  \\
        &  & \\
        & 2 \ar@{.}[d] \ar@{.}[ddl]   \ar@{-}[ddr]     &  \\
        &  \ar@{.}[dl]   \ar@{-}[dr] & \\
        b \ar@{-}[dr] & & b \ar@{-}[dl]\\
        & a 
      }
    }
    \label{fig:diff1:P}
    \caption{Poset $(P, \leq)$}
  \end{subfigure}
  \begin{subfigure}{0.58\textwidth}
    \xymatrix@C=12pt{
      a \ar[d]_{m_0} &
      a \ar[d]_{k_0} \ar[l]_{\id{a}} \ar[rr]^{(a,0)} & &
      0 \ar@/^.35cm/[drrrr]|(.315)\hole^(.75){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} & & 
      a \ar@/_.35cm/[dllll]_(.75){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)} & &
      a \ar[d]^{(a,0)}\ar@{>->}[ll]_{\id{a}} \ar[rr]^{(a,b)} &&
      b \ar[d]^{(b,0)}\\
      c &
      \ar@{>->}[l]^{\id{c}} c \ar[rrr]_{(c,0)}& & & 0 &&&
      \ar@{>->}[lll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0
    }
    \label{fig:diff1:rew}
    \caption{Two rewriting steps}
  \end{subfigure}
  \begin{subfigure}{0.20\textwidth}
    \xymatrix@C=16pt{a \ar[d]_{(a,c)}& a
      \ar[d]_{(a,c)}\ar@{>->}[l]_{\id{a}} \ar[r]^{(a,b)} & b \\c & c
      \ar[l]^{\id{c}}
    }
    \label{ex:diff1:no-step}
    \caption{No rewriting step}
  \end{subfigure}
  \label{fig:diff1} 
  \caption{Sequential independence does not imply switchability}
\end{figure}

\begin{example}
  \label{ex:diff1}
  Consider the poset $(P, \sqsubseteq)$ in Fig.~\ref{fig:diff1} where
  $P = \mathbb{Z} \cup \{a,b,c\}$ and $\sqsubseteq$ is defined by
  $m \sqsubseteq n$ if $m \geq n$, $a \sqsubseteq x$ for all $x \in P$
  and $b, c \sqsubseteq n$ for all $n \in \mathbb{N}$.
  %
  Let $\X$ be the  
  category associated to such order, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. 
  %
  Consider a rewriting
  system whose set of rules contains the following 
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ in Fig.~\ref{fig:diff1:rew}.
  % \[
  %   \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
  %     \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
  %     \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
  %     \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
  %     \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
  %     \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
  %     \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]
  %
  Note that the two direct derivations are sequential
  independent. However there is no switch since the rule applied by
  $\dder{D}_1$ cannot be applied to $c$. In fact, the unique morphism
  $a\to c$ is $(a,c)$, thus we get the diagram in Fig.~\ref{ex:diff1:no-step}. But $b$ and $c$
  do not have a supremum in the poset underlying $\X$, thus the arrows
  $(a,b)\colon a\to b$ and $(a,c)\colon a\to c$ do not have a
  pushout. Hence we do not get a direct derivation from $c$.
\end{example}

@@

\begin{definition}
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph where $s,t\colon E\rightrightarrows V$.
  We define the \emph{depth} of a vertex $v\in V$ as
  \[
    \dph(v)=\begin{cases}
      0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
      m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v
      \} 
    \end{cases}
  \]
\end{definition}

@@

\begin{definition}[Node-merging graphical rewriting systems]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph
  and $\G$ the free category over $\mathcal{G}$. The
  category $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category
  $\Set^{\G}$. A
  \emph{node-merging graphical rewriting system} is a left-linear
  rewriting system $(\gph{G}, \R)$ such that  the
  following hold for each rule
  $(l\colon K\to L, r\colon K\to R)$ in $\R$
  \begin{enumerate}
  \item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$
    with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in
    $\mathcal{G}$) such that $x$ belongs to the image of
    $L(p)\colon L(w)\to L(v)$;
  \item $r\colon K\to R$ is mono on vertices of depth $0$, i.e. for every
    $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$
    is mono.
  \end{enumerate}
\end{definition}

>>>>

\begin{figure}
  \begin{subfigure}{0.2\textwidth}
    {\small
      \xymatrix@C=16pt@R=1pt{
        & 0 \ar@{-}[dd] \ar@{-}[ddddddl] \ar@{-}[ddddddr] &  \\
        &  & \\
        & 1 \ar@{-}[dd] \ar@{-}[ddddl]  \ar@{-}[ddddr]   &  \\
        &  & \\
        & 2 \ar@{.}[d] \ar@{.}[ddl]   \ar@{-}[ddr]     &  \\
        &  \ar@{.}[dl]   \ar@{-}[dr] & \\
        b \ar@{-}[dr] & & b \ar@{-}[dl]\\
        & a 
      }
    }
    \caption{Poset $(P, \leq)$}
    \label{fig:diff1:P}
  \end{subfigure}
  \begin{subfigure}{0.58\textwidth}
    \xymatrix@C=12pt{
      a \ar[d]_{m_0} &
      a \ar[d]_{k_0} \ar[l]_{\id{a}} \ar[rr]^{(a,0)} & &
      0 \ar@/^.35cm/[drrrr]|(.315)\hole^(.75){\id{0}}
      \ar[dr]|(.3)\hole_{\id{0}} & & 
      a \ar@/_.35cm/[dllll]_(.75){(a,c)}
      \ar[dl]|(.3)\hole^{(a,0)} & &
      a \ar[d]^{(a,0)}\ar@{>->}[ll]_{\id{a}} \ar[rr]^{(a,b)} &&
      b \ar[d]^{(b,0)}\\
      c &
      \ar@{>->}[l]^{\id{c}} c \ar[rrr]_{(c,0)}& & & 0 &&&
      \ar@{>->}[lll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0
    }
    \caption{Two rewriting steps}
    \label{fig:diff1:rew}
  \end{subfigure}
  \begin{subfigure}{0.20\textwidth}
    \xymatrix@C=16pt{a \ar[d]_{(a,c)}& a
      \ar[d]_{(a,c)}\ar@{>->}[l]_{\id{a}} \ar[r]^{(a,b)} & b \\c & c
      \ar[l]^{\id{c}}
    }
    \caption{No rewriting step}
    \label{ex:diff1:no-step}
  \end{subfigure}
  \caption{Sequential independence does not imply switchability}
  \label{fig:diff1} 
\end{figure}

\begin{example}
  \label{ex:diff1}
  Consider the poset $(P, \sqsubseteq)$ in Fig.~\ref{fig:diff1:P} where
  $P = \mathbb{Z} \cup \{a,b,c\}$ and $\sqsubseteq$ is defined by
  $m \sqsubseteq n$ if $m \geq n$, $a \sqsubseteq x$ for all $x \in P$
  and $b, c \sqsubseteq n$ for all $n \in \mathbb{N}$.
  %
  Let $\X$ be the  
  category associated to such order, which by \Cref{rem:iso} is
  $\mathsf{Iso(\X)}$-adhesive. 
  %
  Consider a rewriting
  system whose set of rules contains the following 
  \[\xymatrix{a & a \ar[r]^{(a, 0)} \ar[l]_{\id{a}} & 0 & a & a
      \ar[r]^{(a, b)} \ar[l]_{\id{a}} & b }\]
  We can then consider the  derivation
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ in Fig.~\ref{fig:diff1:rew}.
  % \[
  %   \xymatrix@C=30pt{a \ar[d]_{m_0}&& a \ar[d]_{k_0}\ar[ll]_{\id{a}}
  %     \ar[r]^{(a,0)} & 0 \ar@/^.35cm/[drrr]|(.315)\hole_(.45){\id{0}}
  %     \ar[dr]|(.3)\hole_{\id{0}} && a \ar@/_.35cm/[dlll]^(.45){(a,c)}
  %     \ar[dl]|(.3)\hole^{(a,0)}& a \ar[d]^{(a,0)}\ar@{>->}[l]_{\id{a}}
  %     \ar[rr]^{(a,b)} && b \ar[d]^{(b,0)}\\c &&
  %     \ar@{>->}[ll]^{\id{c}} c \ar[rr]_{(c,0)}&& 0 &&
  %     \ar@{>->}[ll]^{\id{0}} 0 \ar[rr]_{\id{0}}&& 0}\]
  %
  Note that the two direct derivations are sequential
  independent. However there is no switch since the rule applied by
  $\dder{D}_1$ cannot be applied to $c$. In fact, the unique morphism
  $a\to c$ is $(a,c)$, thus we get the diagram in Fig.~\ref{ex:diff1:no-step}. But $b$ and $c$
  do not have a supremum in the poset underlying $\X$, thus the arrows
  $(a,b)\colon a\to b$ and $(a,c)\colon a\to c$ do not have a
  pushout. Hence we do not get a direct derivation from $c$.
\end{example}

@@

\begin{definition}
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph where $s,t\colon E\rightrightarrows V$.
  We define the \emph{depth} of a vertex $v\in V$ as\todo{ma ci serve tutta la depth?}
  \[
    \dph(v)=\begin{cases}
      0 & \text{there is no $e\in E$ such that $t(e)=v$}\\
      m+1 &\text{where } m=\min\{\dph(s(e)) \mid e\in E \text{ such that } t(e)=v
      \} 
    \end{cases}
  \]
\end{definition}

@@

\begin{definition}[Node-merging graphical rewriting systems]
  Let $\mathcal{G}=(V, E, s, t)$ be a finite directed acyclic graph
  and $\G$ the free category over $\mathcal{G}$. The
  category $\gph{G}$ of \emph{$\mathcal{G}$-graphs} is the category
  $\Set^{\G}$. A
  \emph{node-merging graphical rewriting system}\todo{altro nome? source/leaf preserving} is a left-linear
  rewriting system $(\gph{G}, \R)$ such that  the
  following hold for each rule
  $(l\colon K\to L, r\colon K\to R)$ in $\R$
  \begin{enumerate}
  \item for every $v\in V$ and $x\in L(v)$, there exists a vertex $w$
    with depth $0$ and an arrow $p\colon w\to v$ (i.e.~a path in
    $\mathcal{G}$) such that $x$ belongs to the image of
    $L(p)\colon L(w)\to L(v)$;
  \item $r\colon K\to R$ is mono on vertices of depth $0$, i.e. for every
    $v\in V$ such that $\dph(v)=0$, the component $r_v:K(v)\to R(v)$
    is mono.
  \end{enumerate}
\end{definition}
